Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-08 paper/HH4b.tex
more work polishing the text

\abstract{
  The measurement of Higgs pair production is one of the cornerstones of
  the LHC program for the next years.
  %
  Double Higgs production provides a unique window to the electroweak
  symmetry breaking mechanism due to its sensitivity
  to the Higgs trilinear coupling.
  %
  In this work, we study the feasibility of a measurement of
  Higgs pair production in the
  $b\bar{b}b\bar{b}$ final state at the LHC.
  %
  Our analysis is based on a combination of traditional cut-based
  methods with state-of-the-art
  multivariate techniques.
  %
  We account for  all relevant
  backgrounds, including the contribution from light and charm
  jet mistags that is found to be comparable
  to the irreducible $4b$ QCD background.
  %
  We demonstrate the robustness of our analysis in a high pile-up
  environment.
  %
  For an integrated luminosity of $\mathcal{L}=3$ ab$^{-1}$,
  an overall signal significance of  $S/\sqrt{B}\simeq 4$ is obtained.
  %
  Our analysis indicates
  that the $b\bar{b}b\bar{b}$ channel
alone should allow the observation of double Higgs production
  at the High Luminosity LHC.
}

>>>>

\abstract{
  The measurement of Higgs pair production is one of the cornerstones of
  the LHC program for the next years.
  %
  Double Higgs production provides a crucial window to the electroweak
  symmetry breaking mechanism and has unique sensitivity
  to the Higgs trilinear coupling.
  %
  In this work, we study the feasibility of a measurement of
  Higgs pair production in the
  $b\bar{b}b\bar{b}$ final state at the LHC.
  %
  Our analysis is based on a combination of traditional cut-based
  methods with state-of-the-art
  multivariate techniques.
  %
  We account for  all relevant
  backgrounds, including the contribution from light and charm
  jet mistags, which turns out to be comparable
  to the irreducible $4b$ QCD background.
  %
  We demonstrate the robustness of our analysis strategy in a high pile-up
  environment.
  %
  For an integrated luminosity of $\mathcal{L}=3$ ab$^{-1}$,
  a signal significance of  $S/\sqrt{B}\simeq 4$ is obtained,
  indicating
  that the $b\bar{b}b\bar{b}$ channel
alone should allow for the observation of double Higgs production
  at the High Luminosity LHC.
}

### 2015-12-08 paper/sec-intro.tex
more work polishing the text

The measurement of double Higgs production will be one of the main
physics goals of the LHC program in its recently started high-energy
phase, as well as for its future high-luminosity upgrade (HL-LHC),
which aims to accumulate a total integrated
luminosity of 3 ab$^{-1}$~\cite{ATLAS:2013hta,CMS:2013xfa}.
%
Higgs pair production is directly sensitive to the
Higgs trilinear coupling $\lambda$~\cite{baglio}, and thus 
provides crucial
information on the electroweak symmetry breaking mechanism.
%
It can also probe the underlying strength of the Higgs interactions
at high energies, and can be used to test the composite nature of the 
Higgs boson~\cite{Giudice:2007fh,Contino:2010mh}.
%
While Standard Model (SM) cross-sections are small,
many Beyond the SM (BSM)
scenarios predict enhanced rates for double Higgs production, and
thus searches for this process have already been performed by ATLAS and CMS with Run I data~\cite{Aad:2015uka,Aad:2014yja,Khachatryan:2015yea,Chatrchyan:2011wt}
and will continue at Run II.
%
In addition, Higgs pair production will also be of
crucial relevance at
any future high-energy 
collider, either at a 100 TeV circular machine~\cite{Barr:2014sga,Papaefstathiou:2015iba,Azatov:2015oxa} or at
a linear or circular electron-positron collider~\cite{Contino:2013gna}.

Analogously to single Higgs production~\cite{Dittmaier:2012vm}, 
in the SM the dominant mechanism for the production of a pair of
Higgs bosons at the LHC is 
gluon fusion (see Ref.~\cite{baglio,Frederix:2014hta} and
references therein).
%
For a center-of-mass energy $\sqrt{s} = 14\,$TeV, the
next-to-next-to-leading order (NNLO)
total cross section is approximately $40\,$fb~\cite{deFlorian:2013jea},
which is increased by a further few percent once
next-to-next-to-leading logarithmic
(NNLL) corrections
are accounted for~\cite{deFlorian:2015moa}.
%
Feasibility studies in the case of a SM-like Higgs boson
in gluon fusion have been performed for several different final states, including
$b\bar b\gamma\gamma$~\cite{Baur:2003gp,Barger:2013jfa},
$b\bar{b}\tau^+\tau^-$~\cite{Baur:2003gpa,Barr:2013tda,Dolan:2012rv,Dolan:2013rja},
$b\bar{b}W^+W^-$~\cite{Dolan:2012rv,Papaefstathiou:2012qe} and
$b\bar{b}b\bar{b}$~\cite{Baur:2003gpa,Dolan:2012rv,Wardrope:2014kya,deLima:2014dta,Barger:2013jfa}.
%
While these studies differ in their qualitative conclusions,
a consistent picture emerges 
that the ultimate precision in the determination of the Higgs trilinear
coupling $\lambda$
will only be achieved from the combination of various final states.

In addition to the gluon-fusion production mode, Higgs pairs
can also be produced in the vector-boson fusion
channel $hhjj$~\cite{Contino:2010mh,Dolan:2013rja,Dolan:2015zja,
  Brooijmans:2014eja},
the associated production modes
$hhW$ and $hhZ$~\cite{Barger:1988jk,baglio,Cao:2015oxx}
(also known as Higgs-Strahlung),
and also in association
with top quark pairs $hht\bar{t}$~\cite{Englert:2014uqa}.
%
All these channels are challenging due to the small production
rates: at 14 TeV, the inclusive total cross-sections are
2 fb for VBF $hhjj$~\cite{Liu-Sheng:2014gxa},
0.5 fb for $W(Z)hh$~\cite{baglio}
and 1.0 for $hht\bar{t}$~\cite{Englert:2014uqa}.
%
Concentrating on the SM dynamics of Higgs pair production
in different channels, several studies
have investigate the interplay of various kinematical
distributions towards the optimisation of the
extraction of $\lambda$ from the measured
cross-sections, as well as the improvement
in the signal modeling taking into account
higher-order corrections and theoretical
uncertainties~\cite{Slawinska:2014vpa,Chen:2014xra,Goertz:2013kp,
  Frederix:2014hta,Dawson:2015oha,Maltoni:2014eza,Maierhofer:2013sha,Grigo:2013rya,Grigo:2014jma}.
%

While the SM production rates for Higgs
pairs are small,
and require the full integrated luminosity of the HL-LHC
to be accessed, double Higgs production rates can be substantially
enhanced in a variety of Beyond the SM (BSM) scenarios.
%
Feasibility studies of Higgs pair production in New Physics
models have been performed in a number of different frameworks,
including Effective Field
Theories (EFTs) with higher-dimensional
operators and anomalous values of the
Higgs couplings~\cite{Nishiwaki:2013cma,Dall'Osso:2015aia,Azatov:2015oxa,Liu:2014rba,Goertz:2014qta,He:2015spf,Grober:2015cwa}, resonant production
in models such as extra dimensions~\cite{Gouzevitch:2013qca,Cooper:2013kia,No:2013wsa}, as well as
in the context of Supersymmetry or Two Higgs Doublet models (2HDMs)~\cite{Belyaev:1999kk,Han:2013sga,Hespel:2014sla,Wu:2015nba,Cao:2014kya,Ellwanger:2013ova}.
%
In general, BSM dynamics modify
the kinematic distributions of the Higgs decay products, for
instance boosting the di-Higgs system, and therefore
different analysis strategies would be required in BSM
Higgs pair searches as compared to SM production measurements.

Searches for the production of Higgs pairs
have already been performed at Run I with 8 TeV
by ATLAS in the $b\bar{b}b\bar{b}$~\cite{Aad:2015uka}
and $b\bar{b}\gamma\gamma$~\cite{Aad:2014yja} final states,
and by
CMS in the same $b\bar{b}b\bar{b}$~\cite{Khachatryan:2015yea}
and $b\bar{b}\gamma\gamma$~\cite{Chatrchyan:2011wt} final
states.
%
%
In addition, ATLAS has presented~\cite{Aad:2015xja} a combination
of its di-Higgs searches in the $bb\tau\tau,$
$\gamma\gamma WW^*$, $\gamma\gamma bb$, and $bbbb$ final states.
%
Many other exotic searches involves Higgs pairs in the final
state, such as the recent
CMS searches for a heavy Higgs boson $H$~\cite{Khachatryan:2015tha}.

In the context of SM production,
the main advantage of the $b\bar{b}b\bar{b}$ final state is the
enhancement of the signal yield
from the large branching fraction of Higgs bosons into bottom-anti-bottom
pairs, ${\rm BR}\lp H\to b\bar{b}\rp\simeq 0.57$~\cite{Dittmaier:2012vm}.
%
However, a measurement in this channel
needs to deal with an overwhelming QCD multi-jet background.
%
Recent studies of Higgs pair production in this
final state~\cite{Wardrope:2014kya,deLima:2014dta}
estimate that, for an integrated
luminosity of
$\mathcal{L}_{\rm int}$=3000 fb$^{-1}$,
a signal significance of around $S/\sqrt{B}\sim 2.0$ could be obtained.
%
While irreducible backgrounds, such as $4b$ and
$t\bar{t}$, are included in these analysis, the contribution
from reducible processes, such as $bbjj$ and
$jjjj$, which can fake the signal when 
light jets are mistagged as $b$-jets, are not accounted for.
%
As will show, this approximation is not justified since, both due to
combinatorics and to $b$-quark radiation in the
parton shower, the
contribution of the $2b2j$ and $4j$ backgrounds can become comparable or even larger
than the irreducible $4b$ component.

In this work, we revisit the feasibility of SM Higgs pair production by gluon-fusion
in the $b\bar{b}b\bar{b}$ final state.
%
 Our strategy is based on a combination of traditional cut-based
 methods with multivariate analysis (MVA),
 leading to a substantial
  enhancement of the signal discrimination.
  %
  We account for  all relevant
  backgrounds, including the contribution from mistags of
  light and charm jets.
  %
  We also model the impact of the pile-up (PU) effects expected at the
  HL-LHC by means of the {\tt SoftKiller} PU subtraction
  method~\cite{Cacciari:2014gra}.
  %
  By combining the contribution from
 the different event categories
  we achieve a total signal significance of $S/\sqrt{B}\simeq 4.0$
  for $\mathcal{L}_{\rm int}$=3000 fb$^{-1}$.
%
  These results emphasize the crucial importance that the $b\bar{b}b\bar{b}$ final state
  has for studies of Higgs pair production and for the determination of
  $\lambda$, as well as the advantages of the use of modern statistical
  analysis methods.

The structure of this paper is the following.
%
In Sect.~\ref{mcgeneration} we discuss the modeling of the signal
and background processes using Monte Carlo event generators.
%
In Sect.~\ref{sec:analysis}
we introduce our analysis strategy, and in particular
the classification of each individual event into
different categories according to their topology.
%
The results for the traditional cut-based analysis
are presented in Sect.~\ref{sec:results}.
%
In Sect.~\ref{sec:mva} we show how we can enhance signal
significance using multivariate techniques, and
illustrate the robustness of our results under realistic pile-up
conditions expected at the HL-LHC.
%
Finally, in Sect.~\ref{sec:conclusions} we conclude and outline
how our results can be extended to estimate the accuracy
in the determination of trilinear coupling $\lambda$ and the constraints on
BSM scenarios.

>>>>

The measurement of double Higgs production will be one of the central
physics goals of the LHC program in its recently started high-energy
phase, as well as for its future high-luminosity upgrade (HL-LHC),
which aims to accumulate a total integrated
luminosity of 3 ab$^{-1}$~\cite{ATLAS:2013hta,CMS:2013xfa}.
%
Higgs pair production is directly sensitive to the
Higgs trilinear coupling $\lambda$~\cite{baglio} and 
provides crucial
information on the electroweak symmetry breaking mechanism.
%
It also probes the underlying strength of the Higgs interactions
at high energies, and can be used to test the composite nature of the 
Higgs boson~\cite{Giudice:2007fh,Contino:2010mh}.
%
While Standard Model (SM) cross-sections are small,
many Beyond the SM (BSM)
scenarios predict enhanced rates for double Higgs production, and
thus searches for this process have already been performed by ATLAS and CMS with Run I data~\cite{Aad:2015xja,Aad:2015uka,Aad:2014yja,Khachatryan:2015yea,Chatrchyan:2011wt}
and will continue at Run II.
%
In addition, Higgs pair production will also be of
crucial relevance at
any future high-energy 
collider, either at a 100 TeV circular machine~\cite{Arkani-Hamed:2015vfh,Barr:2014sga,Papaefstathiou:2015iba,Azatov:2015oxa} or at
a linear or circular electron-positron collider~\cite{Contino:2013gna}.

Analogously to single Higgs production~\cite{Dittmaier:2012vm}, 
in the SM the dominant mechanism for the production of a pair of
Higgs bosons at the LHC is 
gluon fusion (see Refs.~\cite{baglio,Frederix:2014hta} and
references therein).
%
For a center-of-mass energy $\sqrt{s} = 14\,$TeV, the
next-to-next-to-leading order (NNLO)
total cross section is approximately $40\,$fb~\cite{deFlorian:2013jea},
which is increased by a further few percent once
next-to-next-to-leading logarithmic
(NNLL) corrections
are accounted for~\cite{deFlorian:2015moa}.
%
Feasibility studies in the case of a SM-like Higgs boson
in gluon fusion at the LHC have been performed for different final states, including
$b\bar b\gamma\gamma$~\cite{Baur:2003gp,Barger:2013jfa},
$b\bar{b}\tau^+\tau^-$~\cite{Baur:2003gpa,Barr:2013tda,Dolan:2012rv,Dolan:2013rja},
$b\bar{b}W^+W^-$~\cite{Dolan:2012rv,Papaefstathiou:2012qe} and
$b\bar{b}b\bar{b}$~\cite{Baur:2003gpa,Dolan:2012rv,Wardrope:2014kya,deLima:2014dta,Barger:2013jfa}.
%
While these studies differ in their quantitative conclusions,
a consistent picture emerges 
that the ultimate precision in the determination of the Higgs trilinear
coupling $\lambda$ requires the full integrated luminosity
of the HL-LHC, $\mathcal{L}=3$ ab$^{-1}$,
and will rely on the combination of all relevant final states.

In addition to the gluon-fusion production mode, Higgs pairs
can also be produced in the vector-boson fusion
channel $hhjj$~\cite{Contino:2010mh,Dolan:2013rja,Dolan:2015zja,
  Brooijmans:2014eja},
the associated production modes
$hhW$ and $hhZ$~\cite{Barger:1988jk,baglio,Cao:2015oxx}
(also known as Higgs-Strahlung),
and also in association
with top quark pairs $hht\bar{t}$~\cite{Englert:2014uqa}.
%
All these channels are challenging due to the small production
rates: at 14 TeV, the inclusive total cross-sections are
2.0 fb for VBF $hhjj$~\cite{Liu-Sheng:2014gxa},
0.5 fb for $W(Z)hh$~\cite{baglio}
and 1.0 for $hht\bar{t}$~\cite{Englert:2014uqa}.
%
The interplay between  kinematic
distributions for the
extraction of $\lambda$ from the measured
cross-sections, and the role of the associated theoretical
uncertainties, have been intensely scrutinized
recently~\cite{Slawinska:2014vpa,Chen:2014xra,Goertz:2013kp,
  Frederix:2014hta,Dawson:2015oha,Maltoni:2014eza,Maierhofer:2013sha,Grigo:2013rya,Grigo:2014jma}.
%

While the SM production rates for Higgs
pairs are small, these can be substantially
enhanced in a variety of BSM scenarios.
%
Feasibility studies of Higgs pair production in New Physics
models have been performed in a number of different frameworks,
including Effective Field
Theories (EFTs) with higher-dimensional
operators and anomalous 
Higgs couplings~\cite{Nishiwaki:2013cma,Dall'Osso:2015aia,Azatov:2015oxa,Liu:2014rba,Goertz:2014qta,He:2015spf,Grober:2015cwa}, resonant production
in models such as extra dimensions~\cite{Gouzevitch:2013qca,Cooper:2013kia,No:2013wsa}, or Supersymmetry and
Two Higgs Doublet models (2HDMs)~\cite{Belyaev:1999kk,Han:2013sga,Hespel:2014sla,Wu:2015nba,Cao:2014kya,Ellwanger:2013ova}.
%
Since BSM dynamics modify
the kinematic distributions of the Higgs decay products, for
instance boosting the di-Higgs system,
different analysis strategies  might be required for BSM
Higgs pair searches as compared to SM measurements.

Searches for the production of Higgs pairs
have already been performed at Run I with 8 TeV
by ATLAS in the $b\bar{b}b\bar{b}$~\cite{Aad:2015uka}
and $b\bar{b}\gamma\gamma$~\cite{Aad:2014yja} final states,
and by
CMS in the same $b\bar{b}b\bar{b}$~\cite{Khachatryan:2015yea}
and $b\bar{b}\gamma\gamma$~\cite{Chatrchyan:2011wt} final
states.
%
%
In addition, ATLAS has presented~\cite{Aad:2015xja} a combination
of its di-Higgs searches in the $bb\tau\tau,$
$\gamma\gamma WW^*$, $\gamma\gamma bb$, and $bbbb$ final states.
%
Many other exotic searches involves Higgs pairs in the final
state, such as the recent
searches for heavy Higgs bosons $H$~\cite{Khachatryan:2015tha}.

In the context of SM production,
the main advantage of the $b\bar{b}b\bar{b}$ final state is the
enhancement of the signal yield
from the large branching fraction of Higgs bosons into bottom-anti-bottom
pairs, ${\rm BR}\lp H\to b\bar{b}\rp\simeq 0.57$~\cite{Dittmaier:2012vm}.
%
However, a measurement in this channel
needs to deal with an overwhelming QCD multi-jet background.
%
Recent studies of Higgs pair production in this
final state~\cite{Wardrope:2014kya,deLima:2014dta}
estimate that, for an integrated
luminosity of
$\mathcal{L}$=3000 fb$^{-1}$,
a signal significance of around $S/\sqrt{B}\sim 2.0$ could be obtained.
%
While irreducible backgrounds, such as $4b$ and
$t\bar{t}$, are included in these analysis, the contribution
from reducible processes, such as $bbjj$ and
$jjjj$, which can contribute to the signal yield when 
light jets are mis-identified as $b$-jets,
are not accounted for.
%
As will show, due to both
selection effects and $b$-quark radiation in the
parton shower, the
contribution of the $2b2j$ process is actually comparable to
the irreducible $4b$ component.

In this work, we revisit the feasibility of SM Higgs pair production in
gluon-fusion
in the $b\bar{b}b\bar{b}$ final state at the LHC.
%
 Our strategy is based on a combination of traditional cut-based
 methods with multivariate analysis (MVA).
  %
  We account for  all relevant
  backgrounds, including the contribution from mis-identified
  light and charm jets.
  %
  We also assess the robustness of our analysis strategy in
  an environment with high pile-up (PU).
  %
  We find an overall signal significance of $S/\sqrt{B}\simeq 4.0$
  for a total integrated
  luminosity of $\mathcal{L}$=3000 fb$^{-1}$,
   indicating
  that the $b\bar{b}b\bar{b}$ channel
alone should allow for the observation of double Higgs production
  at the HL-LHC.
%

The structure of this paper is the following.
%
In Sect.~\ref{mcgeneration} we present the modeling of the signal
and background processes with Monte Carlo event generators.
%
In Sect.~\ref{sec:analysis}
we introduce our analysis strategy, and in particular
the classification of individual events into
different categories according to their topology.
%
The results the cut-based analysis
are then presented in Sect.~\ref{sec:results}.
%
In Sect.~\ref{sec:mva} we illustrate the enhancement of signal
significance using multivariate techniques, and
assess the robustness of our results in a high-PU environment.
%
In Sect.~\ref{sec:conclusions} we conclude and outline
future studies to extend our results to estimate the accuracy
in the determination of trilinear coupling $\lambda$, as well as
to provide
 constraints in
BSM scenarios.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-12-08 paper/sec-mva.tex
more work polishing the text

\subsection{MVA strategy: deep neural networks}

@@

 In order to avoid overfitting,
 we have used a cross-validation stopping
 criterion, in particular, the same one as
 that used in the NNPDF3.0 analysis~\cite{Ball:2014uwa}
 does not modify the results.
 %
 This cross-validation entails dividing the input MC dataset into two disjoint sets,
 and use one for the training and the other for the validation: the optimal
 stopping point is then indicated by the minimum of the error function
 Eq.~(\ref{cross-entropy}) for the validation sub-sample.

 \subsection{Input kinematical variables for the MVA}
 \label{sec:input}

@@

\subsection{Signal significance after the MVA}
\label{sec:signalsignificance}

@@

\subsection{MVA revisited in the presence of PU}

>>>>

\subsection{Deep artificial neural networks}

@@

 In order to avoid over-fitting,
 we have used a cross-validation stopping
 criterion, in particular, the same one as
 that used in the NNPDF3.0 analysis~\cite{Ball:2014uwa}
 does not modify the results.
 %
 This cross-validation entails dividing the input MC dataset into two disjoint sets,
 and use one for the training and the other for the validation: the optimal
 stopping point is then indicated by the minimum of the error function
 Eq.~(\ref{cross-entropy}) for the validation sub-sample.

 \subsection{Input kinematical variables}
 \label{sec:input}

@@

\subsection{Post-MVA signal significance}
\label{sec:signalsignificance}

@@

\subsection{Impact of PU}
